Reanalysis Step 1: Data Preparation

### 2025-08-08 reviewer-response.Rmd

---
title: "Reviewer Response Analysis"
author: "Erik Westlund"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

@@

# Effect for PDR (main + interaction)
if("ever_lapse_binary:dr_severityPDR" %in% names(coef_vals)) {
  manual_effects$Log_Odds[3] <- coef_vals["ever_lapse_binary"] + 
                                 coef_vals["ever_lapse_binary:dr_severityPDR"]
  # SE calculation for sum of coefficients
  manual_effects$SE[3] <- sqrt(
    vcov_mat["ever_lapse_binary", "ever_lapse_binary"] +
    vcov_mat["ever_lapse_binary:dr_severityPDR", "ever_lapse_binary:dr_severityPDR"] +
    2 * vcov_mat["ever_lapse_binary", "ever_lapse_binary:dr_severityPDR"]
  )
}

# Calculate CIs and OR
manual_effects <- manual_effects |>
  mutate(
    CI_Lower = Log_Odds - 1.96 * SE,
    CI_Upper = Log_Odds + 1.96 * SE,
    Odds_Ratio = exp(Log_Odds),
    # Calculate p-values (two-sided test)
    Z_stat = Log_Odds / SE,
    P_Value = 2 * pnorm(-abs(Z_stat))
  ) |>
  mutate(
    Log_Odds = round(Log_Odds, 4),
    SE = round(SE, 4),
    CI_Lower = round(CI_Lower, 4),
    CI_Upper = round(CI_Upper, 4),
    Odds_Ratio = round(Odds_Ratio, 4),
    P_Value = format.pval(P_Value, digits = 4, eps = 0.0001)
  ) |>
  select(-Z_stat)

knitr::kable(manual_effects,
             caption = "Effect of Lapsing Within Each DR Severity Group (Manual Calculation)",
             row.names = FALSE)
```

## Visualization of Interaction Effects

```{r visualize_interactions, fig.height=6, fig.width=10}
# Use the manual_effects data for visualization
plot_data <- manual_effects

# Forest plot - Odds ratios only
p_forest <- ggplot(plot_data, aes(x = DR_Severity, y = Odds_Ratio)) +
  geom_point(size = 4, color = "darkred") +
  geom_errorbar(aes(ymin = exp(CI_Lower), ymax = exp(CI_Upper)), 
                width = 0.2, size = 1, color = "darkred") +
  geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.5) +
  labs(title = "Odds Ratio of Vision Impairment for Lapsers vs Non-Lapsers",
       subtitle = "By DR severity with 95% confidence intervals",
       x = "DR Severity",
       y = "Odds Ratio") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12)) +
  scale_y_continuous(trans = "log2", 
                     breaks = c(0.5, 1, 2, 4, 8),
                     labels = c("0.5", "1", "2", "4", "8"))

print(p_forest)

# Bar chart showing magnitude of effects with error bars
p_magnitude <- ggplot(plot_data, aes(x = DR_Severity, y = Log_Odds, fill = DR_Severity)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper), 
                width = 0.2, size = 1) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
  labs(title = "Magnitude of Lapse Effect Across DR Severity",
       subtitle = "Note: Confidence intervals for No DR and NPDR overlap",
       x = "DR Severity",
       y = "Log Odds (Effect Size)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        legend.position = "none") +
  scale_fill_manual(values = c("No_DR" = "#1f77b4", 
                               "NPDR" = "#ff7f0e", 
                               "PDR" = "#d62728")) +
  geom_text(aes(label = sprintf("OR: %.2f", Odds_Ratio)),
            vjust = ifelse(plot_data$Log_Odds > 0, -1, 1.5),
            size = 4)

print(p_magnitude)
```

## Average Marginal Effects Visualization

```{r marginal_effects_viz, fig.height=5, fig.width=8}
# Calculate average marginal effects using margins or emmeans
# This shows the actual effect of lapsing within each group
ame_plot <- ggplot(plot_data, aes(x = DR_Severity, y = Log_Odds)) +
  geom_point(size = 5, color = "darkblue") +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper), 
                width = 0.15, size = 1.2, color = "darkblue") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5, color = "gray50") +
  labs(title = "Average Marginal Effect of Lapsing by DR Severity",
       subtitle = "Effect on log odds of vision impairment with 95% CIs",
       x = "DR Severity",
       y = "Average Marginal Effect (Log Odds)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12)) +
  # Add annotations for sample sizes
  annotate("text", x = 1, y = min(plot_data$CI_Lower) - 0.1, 
           label = "n=16,130", size = 3.5, color = "gray50") +
  annotate("text", x = 2, y = min(plot_data$CI_Lower) - 0.1, 
           label = "n=5,695", size = 3.5, color = "gray50") +
  annotate("text", x = 3, y = min(plot_data$CI_Lower) - 0.1, 
           label = "n=3,187", size = 3.5, color = "gray50")

print(ame_plot)
```

@@

# Interaction terms
for(term in dr_interaction_terms) {
  if(term %in% rownames(all_coefs)) {
    dr_level <- gsub("ever_lapse_binary:dr_severity", "", term)
    interp_data <- rbind(interp_data, data.frame(
      Term = term,
      Estimate = round(all_coefs[term, "Estimate"], 4),
      Interpretation = paste("Additional effect of lapsing in", dr_level, "vs No DR")
    ))
  }
}

knitr::kable(interp_data,
             caption = "Interpretation of Model Coefficients",
             row.names = FALSE)
```

# Summary

## Key Findings

A formal interaction analysis provides evidence that **==he effect of lapsing from care on vision impairment depends significantly on patients' baseline DR severity (joint interaction test p < 0.05).

### Substantive Results:

1. **Patients with No DR (n=16,130)**: Lapsing from care increases the odds of vision impairment by approximately 17% (OR = 1.17, log odds = 0.160). This effect is statistically significant.

2. **Patients with NPDR (n=5,695)**: Lapsing has a substantially stronger effect, increasing the odds of vision impairment by approximately 47% (OR = 1.47, log odds = 0.386). This effect is also statistically significant and notably larger than in the No DR group (although the error bars do cross over)

3. **Patients with PDR (n=3,187)**: Lapsing shows minimal effect on vision impairment risk (OR ≈ 1.01, log odds = 0.006), and this effect is not statistically significant.

### Response to Reviewer:

This analysis directly addresses the reviewer's concern about "inadequate subgroup testing" by:

- Providing formal statistical tests (joint Wald test) rather than relying on visual overlap alone
- Calculating group-specific effects with proper confidence intervals
- Demonstrating that the interaction is statistically significant (p < 0.05)
- Showing that subgroup effects are meaningfully different, not just statistically different

The significant interaction confirms that a "one-size-fits-all" interpretation of lapsing effects would be misleading. The surprising finding that PDR patients show essentially no harm from lapsing (OR ≈ 1.01) suggests we may need to **reconsider our conceptual understanding of lapsing in advanced disease**. While NPDR patients show substantial risk from care disruption, the null effect in PDR raises questions about whether our retention efforts should be differentially targeted based on disease severity.

### 2025-08-08 reviewer-response.Rmd

```{r visualize_interactions, fig.height=6, fig.width=10}
# Use the manual_effects data for visualization
plot_data <- manual_effects

# Bar chart showing magnitude of effects with error bars
p_magnitude <- ggplot(plot_data, aes(x = DR_Severity, y = Log_Odds, fill = DR_Severity)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper), 
                width = 0.2, size = 1) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
  labs(title = "Magnitude of Lapse Effect Across DR Severity",
       subtitle = "Note: Confidence intervals for No DR and NPDR overlap",
       x = "DR Severity",
       y = "Log Odds (Effect Size)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        legend.position = "none") +
  scale_fill_manual(values = c("No_DR" = "#1f77b4", 
                               "NPDR" = "#ff7f0e", 
                               "PDR" = "#d62728")) +
  geom_text(aes(label = sprintf("OR: %.2f", Odds_Ratio)),
            vjust = ifelse(plot_data$Log_Odds > 0, -1, 1.5),
            size = 4)

print(p_magnitude)
```

## Average Marginal Effects Visualization (Probability Scale)

```{r marginal_effects_viz, fig.height=5, fig.width=8}
# Convert log odds AMEs to probability differences
# Using the delta method approximation or simulation
# For interpretability, we'll calculate the average marginal effect on the probability scale

# Get baseline predicted probabilities (at means of covariates)
# For a rough approximation, use the intercept to get baseline risk
intercept <- coef(model_dr_interaction)["(Intercept)"]
baseline_prob <- exp(intercept) / (1 + exp(intercept))

# Convert log odds effects to approximate probability differences
# Using the derivative of the logistic function at the baseline probability
# d(prob)/d(log_odds) = prob * (1 - prob)
prob_scale_factor <- baseline_prob * (1 - baseline_prob)

# Create probability scale effects
plot_data_prob <- plot_data |>
  mutate(
    AME_Prob = Log_Odds * prob_scale_factor,
    CI_Lower_Prob = CI_Lower * prob_scale_factor,
    CI_Upper_Prob = CI_Upper * prob_scale_factor,
    # Convert to percentage points for easier interpretation
    AME_Pct = AME_Prob * 100,
    CI_Lower_Pct = CI_Lower_Prob * 100,
    CI_Upper_Pct = CI_Upper_Prob * 100
  )

# Plot average marginal effects on probability scale
ame_plot <- ggplot(plot_data_prob, aes(x = DR_Severity, y = AME_Pct)) +
  geom_point(size = 5, color = "darkblue") +
  geom_errorbar(aes(ymin = CI_Lower_Pct, ymax = CI_Upper_Pct), 
                width = 0.15, size = 1.2, color = "darkblue") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5, color = "gray50") +
  labs(title = "Average Marginal Effect of Lapsing by DR Severity",
       subtitle = "Effect on probability of vision impairment (percentage points) with 95% CIs",
       x = "DR Severity",
       y = "Average Marginal Effect (Percentage Points)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12)) +
  # Add annotations for sample sizes
  annotate("text", x = 1, y = min(plot_data_prob$CI_Lower_Pct) - 1, 
           label = "n=16,130", size = 3.5, color = "gray50") +
  annotate("text", x = 2, y = min(plot_data_prob$CI_Lower_Pct) - 1, 
           label = "n=5,695", size = 3.5, color = "gray50") +
  annotate("text", x = 3, y = min(plot_data_prob$CI_Lower_Pct) - 1, 
           label = "n=3,187", size = 3.5, color = "gray50")

print(ame_plot)

# Also display the probability scale effects in a table
prob_effects_table <- plot_data_prob |>
  select(DR_Severity, AME_Pct, CI_Lower_Pct, CI_Upper_Pct) |>
  mutate(
    AME_Pct = round(AME_Pct, 2),
    CI_Lower_Pct = round(CI_Lower_Pct, 2),
    CI_Upper_Pct = round(CI_Upper_Pct, 2),
    Effect = paste0(AME_Pct, "% (", CI_Lower_Pct, "%, ", CI_Upper_Pct, "%)")
  ) |>
  select(DR_Severity, Effect)

knitr::kable(prob_effects_table,
             caption = "Average Marginal Effect of Lapsing (Percentage Point Change in Probability)",
             col.names = c("DR Severity", "Effect (95% CI)"),
             row.names = FALSE)
```

### 2025-08-08 reviewer-response.Rmd

# Create clean results table
lapse_effects_table <- data.frame(
  DR_Severity = lapse_effects$dr_severity,
  Log_Odds = round(lapse_effects$estimate, 4),
  SE = round(lapse_effects$SE, 4),
  CI_Lower = round(lapse_effects$estimate - 1.96 * lapse_effects$SE, 4),
  CI_Upper = round(lapse_effects$estimate + 1.96 * lapse_effects$SE, 4),
  Odds_Ratio = round(exp(lapse_effects$estimate), 4),
  P_Value = format.pval(lapse_effects$p.value, digits = 4, eps = 0.0001)
)

knitr::kable(lapse_effects_table,
             caption = "Effect of Lapsing Within Each DR Severity Group",
             row.names = FALSE)
```

## Alternative: Manual Calculation of Lapse Effects

```{r manual_lapse_effects}
# Note: This manual calculation should produce identical results to emmeans above
# If emmeans still has issues, we can calculate the effects manually
# Extract coefficients and variance-covariance matrix
coef_vals <- coef(model_dr_interaction)
vcov_mat <- vcov(model_dr_interaction)

@@

# Effect for NPDR (main + interaction)
if("ever_lapse_binary:dr_severityNPDR" %in% names(coef_vals)) {
  manual_effects$Log_Odds[2] <- coef_vals["ever_lapse_binary"] + 
                                 coef_vals["ever_lapse_binary:dr_severityNPDR"]
  # SE calculation for sum of coefficients
  manual_effects$SE[2] <- sqrt(
    vcov_mat["ever_lapse_binary", "ever_lapse_binary"] +
    vcov_mat["ever_lapse_binary:dr_severityNPDR", "ever_lapse_binary:dr_severityNPDR"] +
    2 * vcov_mat["ever_lapse_binary", "ever_lapse_binary:dr_severityNPDR"]
  )
}

# Effect for PDR (main + interaction)
if("ever_lapse_binary:dr_severityPDR" %in% names(coef_vals)) {
  manual_effects$Log_Odds[3] <- coef_vals["ever_lapse_binary"] + 
                                 coef_vals["ever_lapse_binary:dr_severityPDR"]
  # SE calculation for sum of coefficients
  manual_effects$SE[3] <- sqrt(
    vcov_mat["ever_lapse_binary", "ever_lapse_binary"] +
    vcov_mat["ever_lapse_binary:dr_severityPDR", "ever_lapse_binary:dr_severityPDR"] +
    2 * vcov_mat["ever_lapse_binary", "ever_lapse_binary:dr_severityPDR"]
  )
}

# Calculate CIs and OR
manual_effects <- manual_effects |>
  mutate(
    CI_Lower = Log_Odds - 1.96 * SE,
    CI_Upper = Log_Odds + 1.96 * SE,
    Odds_Ratio = exp(Log_Odds),
    # Calculate p-values (two-sided test)
    Z_stat = Log_Odds / SE,
    P_Value = 2 * pnorm(-abs(Z_stat))
  ) |>
  mutate(
    Log_Odds = round(Log_Odds, 4),
    SE = round(SE, 4),
    CI_Lower = round(CI_Lower, 4),
    CI_Upper = round(CI_Upper, 4),
    Odds_Ratio = round(Odds_Ratio, 4),
    P_Value = format.pval(P_Value, digits = 4, eps = 0.0001)
  ) |>
  select(-Z_stat)

knitr::kable(manual_effects,
             caption = "Effect of Lapsing Within Each DR Severity Group (Manual Calculation)",
             row.names = FALSE)

# Verify that emmeans and manual calculations match
if(exists("lapse_effects_table") && nrow(lapse_effects_table) > 0) {
  # Compare the two methods
  comparison <- data.frame(
    DR_Severity = manual_effects$DR_Severity,
    Emmeans_LogOdds = lapse_effects_table$Log_Odds,
    Manual_LogOdds = manual_effects$Log_Odds,
    Difference = abs(lapse_effects_table$Log_Odds - manual_effects$Log_Odds)
  )

  if(max(comparison$Difference) < 0.001) {
    cat("\n✓ Verification: Emmeans and manual calculations produce identical results (differences < 0.001)\n")
  } else {
    cat("\n⚠ Warning: Some differences between emmeans and manual calculations:\n")
    print(comparison)
  }
} else {
  cat("\nNote: Using manual calculation method as emmeans encountered issues with reference grid size.\n")
}
```

@@

```{r visualize_interactions, fig.height=6, fig.width=10}
# Use the manual_effects data for visualization
plot_data <- manual_effects

# Bar chart showing magnitude of effects with error bars
p_magnitude <- ggplot(plot_data, aes(x = DR_Severity, y = Log_Odds, fill = DR_Severity)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper), 
                width = 0.2, size = 1) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
  labs(title = "Magnitude of Lapse Effect Across DR Severity",
       subtitle = "Note: Confidence intervals for No DR and NPDR overlap",
       x = "DR Severity",
       y = "Log Odds (Effect Size)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        legend.position = "none") +
  scale_fill_manual(values = c("No_DR" = "#1f77b4", 
                               "NPDR" = "#ff7f0e", 
                               "PDR" = "#d62728")) +
  geom_text(aes(label = sprintf("OR: %.2f", Odds_Ratio)),
            vjust = ifelse(plot_data$Log_Odds > 0, -1, 1.5),
            size = 4)

print(p_magnitude)
```

## Probability Difference by Lapse Status

```{r probability_differences, fig.height=5, fig.width=8}
# Calculate the actual probabilities for lapsers vs non-lapsers
# This shows the DIFFERENCE IN PROBABILITY of vision impairment between groups

# Get baseline predicted probabilities (at means of covariates)
intercept <- coef(model_dr_interaction)["(Intercept)"]
baseline_prob <- exp(intercept) / (1 + exp(intercept))

# Convert log odds effects to probability differences
# Using the derivative of the logistic function
prob_scale_factor <- baseline_prob * (1 - baseline_prob)

# Create probability scale effects
plot_data_prob <- plot_data |>
  mutate(
    # This is the DIFFERENCE in probability between lapsers and non-lapsers
    Prob_Difference = Log_Odds * prob_scale_factor,
    CI_Lower_Prob = CI_Lower * prob_scale_factor,
    CI_Upper_Prob = CI_Upper * prob_scale_factor
  )

# Plot the probability differences
prob_diff_plot <- ggplot(plot_data_prob, aes(x = DR_Severity, y = Prob_Difference)) +
  geom_point(size = 5, color = "darkblue") +
  geom_errorbar(aes(ymin = CI_Lower_Prob, ymax = CI_Upper_Prob), 
                width = 0.15, size = 1.2, color = "darkblue") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5, color = "gray50") +
  labs(title = "Difference in Probability of Vision Impairment: Lapsers vs Non-Lapsers",
       subtitle = "Average marginal effect of lapsing on probability scale with 95% CIs",
       x = "DR Severity",
       y = "Probability Difference") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12)) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001)) +
  # Add annotations for sample sizes
  annotate("text", x = 1, y = min(plot_data_prob$CI_Lower_Prob) - 0.01, 
           label = "n=16,130", size = 3.5, color = "gray50") +
  annotate("text", x = 2, y = min(plot_data_prob$CI_Lower_Prob) - 0.01, 
           label = "n=5,695", size = 3.5, color = "gray50") +
  annotate("text", x = 3, y = min(plot_data_prob$CI_Lower_Prob) - 0.01, 
           label = "n=3,187", size = 3.5, color = "gray50")

print(prob_diff_plot)

# Display the probability differences in a table
prob_diff_table <- plot_data_prob |>
  mutate(
    Prob_Diff_Pct = sprintf("%.1f%%", Prob_Difference * 100),
    CI_Lower_Pct = sprintf("%.1f%%", CI_Lower_Prob * 100),
    CI_Upper_Pct = sprintf("%.1f%%", CI_Upper_Prob * 100),
    Effect = paste0(Prob_Diff_Pct, " (", CI_Lower_Pct, ", ", CI_Upper_Pct, ")")
  ) |>
  select(DR_Severity, Prob_Difference, Effect)

knitr::kable(prob_diff_table[, c("DR_Severity", "Effect")],
             caption = "Difference in Probability of Vision Impairment: Lapsers vs Non-Lapsers",
             col.names = c("DR Severity", "Probability Difference (95% CI)"),
             row.names = FALSE)

# Also show the raw probabilities
cat("\nFor context, the baseline probability of vision impairment (at intercept) is approximately:", 
    sprintf("%.1f%%", baseline_prob * 100), "\n")
```

### 2025-08-08 reviewer-response.Rmd

# Interpret the joint test
if(joint_test$`Pr(>Chisq)`[2] < 0.05) {
  print("Joint test is significant: The effect of lapsing varies by DR severity")
} else {
  print("Joint test is not significant: The effect of lapsing is similar across DR severity levels")
}
```

## Effect of Lapsing Within Each DR Severity Group

```{r lapse_effects_by_dr}

# Method 1: Use emmeans WITHOUT nuisance factors for true marginalization
# Increase reference grid limit to handle the full factorial
emm_options(rg.limit = 200000)  # Increase limit to handle large reference grid

# Calculate marginal means - this will average over all covariate combinations
emm <- emmeans(model_dr_interaction, 
               ~ ever_lapse_binary | dr_severity)

@@

```{r visualize_interactions, fig.height=6, fig.width=10}
# Use the manual_effects data for visualization
plot_data <- manual_effects

# Bar chart showing magnitude of effects with error bars
p_magnitude <- ggplot(plot_data, aes(x = DR_Severity, y = Log_Odds, fill = DR_Severity)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper), 
                width = 0.2, size = 1) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
  labs(title = "Magnitude of Lapse Effect Across DR Severity",
       subtitle = "Note: Confidence intervals for No DR and NPDR overlap",
       x = "DR Severity",
       y = "Log Odds (Effect Size)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        legend.position = "none") +
  scale_fill_manual(values = c("No_DR" = "#1f77b4", 
                               "NPDR" = "#ff7f0e", 
                               "PDR" = "#d62728")) +
  geom_text(aes(label = sprintf("OR: %.2f", Odds_Ratio)),
            vjust = ifelse(plot_data$Log_Odds > 0, -1, 1.5),
            size = 4)

print(p_magnitude)
```

## Predicted Probabilities and Differences by Lapse Status

```{r probability_table_emmeans}
# Use emmeans to get predicted probabilities for each combination
# WITHOUT nuisance specification for TRUE marginalization

# IMPORTANT: How emmeans works without nuisance specification:
# - It creates a reference grid with all unique combinations of covariates
# - Then averages predictions across this grid
# - This gives us the AVERAGE predicted probability across all covariate patterns
# - This IS true marginalization over the covariate distribution

# Increase reference grid limit to handle all combinations
emm_options(rg.limit = 200000)

# Get predicted probabilities on the response scale
# This will average over all covariate combinations
emm_probs <- emmeans(model_dr_interaction, 
                     ~ ever_lapse_binary | dr_severity,
                     type = "response")

# Show what emmeans is doing
cat("EMmeans is averaging over all covariate combinations in the reference grid\n")
cat("Reference grid size information:\n")
print(str(emm_probs@grid))
cat("\n")

# Convert to data frame for manipulation
prob_summary <- as.data.frame(emm_probs)

# Create a clean table with probabilities for each group
prob_table <- data.frame(
  DR_Severity = rep(unique(prob_summary$dr_severity), each = 2),
  Lapse_Status = rep(c("No Lapse", "Lapse"), length(unique(prob_summary$dr_severity))),
  Probability = prob_summary$prob,
  SE = prob_summary$SE,
  Lower_CI = prob_summary$asymp.LCL,
  Upper_CI = prob_summary$asymp.UCL
)

# Calculate differences manually from the emmeans object
prob_contrasts <- contrast(emm_probs, method = "revpairwise", by = "dr_severity")
prob_diff_emmeans <- as.data.frame(prob_contrasts)

# Create final table combining probabilities and differences
final_prob_table <- data.frame(
  DR_Severity = c("No_DR", "NPDR", "PDR"),
  Prob_No_Lapse = prob_table$Probability[prob_table$Lapse_Status == "No Lapse"],
  Prob_Lapse = prob_table$Probability[prob_table$Lapse_Status == "Lapse"],
  Difference = prob_diff_emmeans$estimate,
  CI_Lower = prob_diff_emmeans$asymp.LCL,
  CI_Upper = prob_diff_emmeans$asymp.UCL,
  P_Value = prob_diff_emmeans$p.value
)

# Format the table for display
display_table <- final_prob_table |>
  mutate(
    Prob_No_Lapse = round(Prob_No_Lapse, 3),
    Prob_Lapse = round(Prob_Lapse, 3),
    Difference = round(Difference, 3),
    CI = paste0("(", round(CI_Lower, 3), ", ", round(CI_Upper, 3), ")"),
    P_Value = format.pval(P_Value, digits = 3, eps = 0.001)
  ) |>
  select(DR_Severity, Prob_No_Lapse, Prob_Lapse, Difference, CI, P_Value)

knitr::kable(display_table,
             caption = "Predicted Probability of Vision Impairment by DR Severity and Lapse Status (Averaged Over All Covariate Patterns)",
             col.names = c("DR Severity", "P(VI | No Lapse)", "P(VI | Lapse)", 
                          "Difference", "95% CI", "P-value"),
             row.names = FALSE,
             digits = 3)
```

## Alternative: Manual Verification of Average Marginal Effects

```{r true_marginal_effects}
# If you want TRUE marginalization (averaging over the distribution of covariates),
# we need to calculate predictions for every observation and then average

cat("Calculating average marginal effects by averaging over all observations...\n\n")

# For each DR severity group, calculate the average effect
marginal_effects <- data.frame(
  DR_Severity = c("No_DR", "NPDR", "PDR"),
  Avg_Prob_No_Lapse = numeric(3),
  Avg_Prob_Lapse = numeric(3),
  Difference = numeric(3),
  SE = numeric(3)
)

for(i in 1:3) {
  dr_level <- c("No_DR", "NPDR", "PDR")[i]

  # Get subset of data for this DR severity
  subset_data <- m_data[m_data$dr_severity == dr_level, ]

  # Predict for everyone with lapse = 0
  subset_data_0 <- subset_data
  subset_data_0$ever_lapse_binary <- 0
  pred_0 <- predict(model_dr_interaction, newdata = subset_data_0, type = "response")

  # Predict for everyone with lapse = 1
  subset_data_1 <- subset_data
  subset_data_1$ever_lapse_binary <- 1
  pred_1 <- predict(model_dr_interaction, newdata = subset_data_1, type = "response")

  # Calculate average probabilities
  marginal_effects$Avg_Prob_No_Lapse[i] <- mean(as.numeric(pred_0), na.rm = TRUE)
  marginal_effects$Avg_Prob_Lapse[i] <- mean(as.numeric(pred_1), na.rm = TRUE)
  marginal_effects$Difference[i] <- marginal_effects$Avg_Prob_Lapse[i] - marginal_effects$Avg_Prob_No_Lapse[i]

  # Bootstrap or delta method for SE would go here
  # For now, using a rough approximation
  marginal_effects$SE[i] <- sd(as.numeric(pred_1) - as.numeric(pred_0), na.rm = TRUE) / sqrt(nrow(subset_data))
}

# Format for display
marginal_display <- marginal_effects |>
  mutate(
    Avg_Prob_No_Lapse = round(Avg_Prob_No_Lapse, 3),
    Avg_Prob_Lapse = round(Avg_Prob_Lapse, 3),
    Difference = round(Difference, 3),
    CI_Lower = round(Difference - 1.96 * SE, 3),
    CI_Upper = round(Difference + 1.96 * SE, 3),
    CI = paste0("(", CI_Lower, ", ", CI_Upper, ")")
  ) |>
  select(DR_Severity, Avg_Prob_No_Lapse, Avg_Prob_Lapse, Difference, CI)

knitr::kable(marginal_display,
             caption = "Average Marginal Effects: Predicted Probabilities Averaged Over Observed Covariate Distribution",
             col.names = c("DR Severity", "Avg P(VI | No Lapse)", "Avg P(VI | Lapse)", 
                          "Difference", "95% CI"),
             row.names = FALSE,
             digits = 3)

cat("\nNote: The 'at reference/mean' table shows predictions for a typical patient.")
cat("\nThe 'average marginal effects' table averages predictions across all patients in each DR group.")
cat("\nThe difference represents how these approaches handle heterogeneity in the covariates.\n")
```

## Visualization of Probability Differences

```{r probability_differences_plot, fig.height=5, fig.width=8}
# Use the emmeans-based differences for plotting
plot_data_prob <- final_prob_table

# Plot the probability differences
prob_diff_plot <- ggplot(plot_data_prob, aes(x = DR_Severity, y = Difference)) +
  geom_point(size = 5, color = "darkblue") +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper), 
                width = 0.15, size = 1.2, color = "darkblue") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5, color = "gray50") +
  labs(title = "Difference in Probability of Vision Impairment: Lapsers vs Non-Lapsers",
       subtitle = "Average marginal effect of lapsing on probability scale with 95% CIs",
       x = "DR Severity",
       y = "Probability Difference") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12)) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001)) +
  # Add annotations for sample sizes
  annotate("text", x = 1, y = min(plot_data_prob$CI_Lower) - 0.01, 
           label = "n=16,130", size = 3.5, color = "gray50") +
  annotate("text", x = 2, y = min(plot_data_prob$CI_Lower) - 0.01, 
           label = "n=5,695", size = 3.5, color = "gray50") +
  annotate("text", x = 3, y = min(plot_data_prob$CI_Lower) - 0.01, 
           label = "n=3,187", size = 3.5, color = "gray50")

print(prob_diff_plot)
```

### 2025-08-08 reviewer-response.Rmd

## Effect of Lapsing Within Each DR Severity Group

```{r lapse_effects_by_dr}

# Method 1: Use emmeans WITHOUT nuisance factors for true marginalization
# Increase reference grid limit to handle the full factorial
emm_options(rg.limit = 200000)  # Increase limit to handle large reference grid

# Check for cached emmeans results
emmeans_cache_file <- "ps_results/emmeans_dr_interaction.rds"
if(file.exists(emmeans_cache_file)) {
  cat("Loading cached emmeans results...\n")
  emmeans_results <- readRDS(emmeans_cache_file)
  emm <- emmeans_results$emm
  lapse_contrasts <- emmeans_results$lapse_contrasts
  cat("Cached results loaded successfully.\n\n")
} else {
  cat("Computing emmeans (this may take several minutes)...\n")
  start_time <- Sys.time()

@@

  # Calculate marginal means - this will average over all covariate combinations
  emm <- emmeans(model_dr_interaction, 
                 ~ ever_lapse_binary | dr_severity)

  # Get contrasts: effect of lapse (1 vs 0) within each DR severity
  lapse_contrasts <- contrast(emm, method = "revpairwise", by = "dr_severity")

  # Save to cache
  emmeans_results <- list(
    emm = emm,
    lapse_contrasts = lapse_contrasts,
    computed_at = Sys.time()
  )
  saveRDS(emmeans_results, file = emmeans_cache_file)

  end_time <- Sys.time()
  cat("Emmeans computation completed in", round(difftime(end_time, start_time, units = "mins"), 2), "minutes\n")
  cat("Results cached to:", emmeans_cache_file, "\n\n")
}
